---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

library(RefManageR)
library(kableExtra)

bib <- ReadBib("papers/references.bib")

files <- list.files("papers/pdfs/")
bib_df <- bib2df::bib2df("papers/references.bib")
df <- bib_df[, c("BIBTEXKEY", "AUTHOR")]
df$AUTHOR <- sapply(df$AUTHOR, function(x) x[[1]])
df <- df[order(df$AUTHOR), ]
df$file <- sort(files)

df$Authors <- sapply(df$BIBTEXKEY, function(x) Citet(bib, x))
df$File <- sprintf("[pdf](%s)", file.path("papers/pdfs", df$file)) 
df$Title <- unname(unlist(bib$title))
df$Title <- gsub("\\{|\\}", "", df$Title)
df$Title <- gsub("\\n", "", df$Title)

```

## Information

This page contains the materials (slide, code, extra) used during the Summer School. For general information and official communications see the page [replicability.stat.unipd.it](https://replicability.stat.unipd.it/)

## General

The document [getting-started](extra/getting-started/getting-started.html) contains a tutorial to install all relevant packages for R and other tools. The entire folder can be downloaded from [here](https://github.com/replicability-stat-unipd/replicability-summer-school-2023/archive/refs/heads/main.zip) or can be cloned using `git`.

```bash
git clone git@github.com:replicability-stat-unipd/replicability-summer-school-2023.git
```

Once downloaded, the project can be opened with `R Studio` simply opening the `replicability-summer-school-2023.Rproj` file. Then using the command `devtools::load_all()` all the custom functions within the `R/` folder can be used.

## Functions

Within slides and exercises we use some custom functions developed for the workshop. These functions can be loaded with `devtools::load_all()`:

- `R/utils.R`: general utilities for managing the project (not useful for the workshop)
- `R/utils-meta.R`: utilities for the meta-analysis workshop
- `R/utils-biostat.R`: utilities for the biostatistic workshop
- `R/utils-replication.R`: utilities for the replication methods workshop

## Useful links

- [Shared code editor](https://etherpad.wikimedia.org/p/replicability-summer-school-2023)
- [Github repository](https://github.com/replicability-stat-unipd/replicability-summer-school-2023)

## Slides

The slides are located into the specific folders. Then when relevant `script/` and `objects/` folders contains extra documents and files included in the slides. The dataset used in the slides are contained also into the `data` folder and datasets can be accessed also using the `data()` function.

| Day | Title | Slides | Source |
|:---:|:---:|:---:|:---:|
| 09-18-2023 | **Introduction to the workshops** | [html](00-intro/slides/00-intro.html) | [qmd](00-intro/slides/00-intro.qmd) |
| 09-18-2023 | **Tools for reproducible research** | [html](01-replication-tools/slides/01-replication-tools.html) | [qmd](01-replication-tools/slides/01-replication-tools.qmd) |
| 09-19-2023 | **Meta-Analysis and Multi-Lab Replication studies** | [html](02-meta-analysis/slides/02-meta-analysis.html) | [qmd](02-meta-analysis/slides/02-meta-analysis.qmd) |
| 09-20-2023 | **Exploring Replicability in Biostatistics** | [html](03-biostatistics/slides/03-biostatistics.html) | [qmd](03-biostatistics/slides/03-biostatistics.qmd) |
| 09-21-2023 | **Statistical Methods for Replication Assessment** | [html](04-replication-methods/slides/04-replication-methods.html) | [qmd](04-replication-methods/slides/04-replication-methods.html) |

## Teamwork

For the teamwork you can find the papers in the table below. You can download the entire folder (`papers/pdfs`) from the following [link](https://minhaskamal.github.io/DownGit/#/home?url=https://github.com/replicability-stat-unipd/replicability-summer-school-2023/tree/master/papers).

```{r}
#| results: asis
#| echo: false
df[, c("Authors", "Title", "File")] |> 
    kable(format = "markdown")
```