---
title: "Eye Cue Anti-Social Behaviour Meta-Analysis - Odds & Risk Ratio"
author: "Keith Dear"
date: "10/17/2016"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Code

This is document is provided to enable replication and critique of our analysis.  To follow it users must first install the packages 'METAFOR', 'XLConnec', DT
```{r}
library("userfriendlyscience")
library("DT")
library("metafor")

```
## Create data file
Create a data file (df1) by reading our .csv file and display it in a datatable

```{r}
df1 <- read.csv("/Users/keith/Documents/OneDrive/001-Oxford Working Documents/003 - Literature Reviews/Anti-Social Behaviour & Eye Cue Review and Meta-Analysis/003 - Evolution & Human Behaviour/Revisions/20180305_ASB_for_R_Extended_w_Nettle_etal_KBT_Meleady_Dog.csv")

datatable(df1, options = list(pageLength = 25))
```

...and display it to check the values:

```{r}
df1
```
## Create data table & calculate log odds ratio & variance
Create a data frame (datMA), containing the log odds ratio and the variance:
```{r}


datMA <- escalc(measure="OR",ai=EYES.ASB.YES, bi=EYES.ASB.NO, ci=CTRL.ASB.YES, di=CTRL.ASB.N, data=df1)
View(datMA)
datatable(datMA, options = list(pageLength = 25))
```
# yi = log odds ratio
# vi = variance of log odds ratio
```{r}


datMA <- datMA %>%
  dplyr::mutate(SE = sqrt(vi))

datMA <- datMA %>%
  dplyr::mutate("95% CI lower" = (yi-(1.96*SE)))

datMA <- datMA %>%
  dplyr::mutate("95% CI upper" = (yi+(1.96*SE)))

datMA <- datMA %>%
  dplyr::mutate("Z" = (yi/SE))

datMA <- datMA %>%
  dplyr::mutate(P = exp(-0.717*Z-0.416*Z^2))


datMA
```

```{r}
write.csv(datMA, "Meta-Analysis_Table.csv")
```

##Q-Test
Next we apply Cochrane's Q-test for homogenity of variance and report the results at 'res1':
```{r}
res1 <- rma.peto(ai=EYES.ASB.YES, bi=EYES.ASB.NO, ci=CTRL.ASB.YES, di=CTRL.ASB.N, data=datMA)
res1
```
Since Q is significant we know we must apply a random effects model (although we would have applied this anyway given the heterogenity in study design):
```{r}
res1<-rma(yi, vi, data = datMA)
res1
```
The estimate is our log odds ratio.
```{r}
exp(-0.4311)
```
Our odds ratio is .65 = 35% reduction in the watching eyes condition.

##Influential Case Diagnostics, Funnel & Forest Plots

Next we test for influential cases in our result 1 (res1):

```{r}
influence(res1)

```

...and goodness of fit:

```{r}
fitstats(res1)
```
#create a forest plot:

```{r}
forest(res1)
```
and a funnel plot:
```{r}
funnel(res1)
```

Then we examine our funnel plot of missing cases (evidence of publication bias): 

```{r}

res1tf <- trimfill(res1)
```
Then we recreate the funnel plot using the trim-fill adjusted data to check for missing cases:
```{r}
funnel(res1tf)
```
##Convert Log Odds to Odds Ratio

We exponentiate the log odds ratio, in this case -0.4311, to convert it to an odds ratio (remember to change the figure here to whatever your output shows if it is different to ours):

# Random Effects Model OR
```{r}
exp(-0.4311)
```
# Fixed Effects Model
```{r}
exp(-0.1320)
```
##RISK RATIO

Finally we create a Risk Ratio table (datMA.RR) from our frequencies table:

```{r}
require("metafor")
datMA.RR <-escalc(measure="RR",ai=CTRL.ASB.YES, bi=CTRL.ASB.N, ci=EYES.ASB.YES, di=EYES.ASB.NO, data=df1)
View(datMA.RR)
```

...and calculate the Risk Ratio

```{r}
res2<-rma(yi, vi, data = datMA.RR)
res2
```
# % decrease = (1-RR) x 100
# % decrease =(1-0.2731)x100
# % decreases = (0.7269)x100
# % decrease = 72.69%

Participants in the eyes condition had 72.69% less risk of behaving anti-socially than those in the control.
Interpreted after http://sphweb.bumc.bu.edu/otlt/MPH-Modules/EP/EP713_Association/EP713_Association3.html



Then create plots for export...
```{r}

dpi = 150

png(filename = "trimfillForest.png", width = (500 * (dpi/72)), height = (300 * (dpi/72)), res = dpi, bg = "transparent")

forest(res1)

dev.off()

dpi = 150

png(filename = "trimfillFunnel2.png", width = (500 * (dpi/72)), height = (300 * (dpi/72)), res = dpi, bg = "transparent")

funnel(res1tf)

dev.off()

```
## PET-PEESE Analysis
```{r}
#install.packages("magrittr")

library(magrittr)
library(metafor)
```
 # Loading the code from https://github.com/Joe-Hilgard/PETPEESE/blob/master/PETPEESE_functions.R

# naive meta-analysis ----
```{r}


naive = function(dataset, ...) {
  rma(yi = Z,
      sei = SE,
      data = datMA,
      method = "FE",
      ...)
}
naive()
```
Naive metaanalysis on the standard errors reverses the findings suggesting that there is an increase in ASB when eyes are present when compared with a control.

# Basic PET
```{r}
PET=function(dataset, error = "additive") {
  if (error == "additive") {
    petOut = rma(yi = Z, 
                 sei = SE, 
                 mods = ~SE, 
                 data=datMA,
                 method = "FE")
  }
  if (error == "multiplicative") {
    petOut = lm(Z ~ SE,
                weights = 1/SE,
                data=datMA)
  }
  return(petOut)
}
PET()
```

# Basic PEESE Function
```{r}

PEESE=function(dataset, error = "additive") {
  if (error == "additive") {
    peeseOut = rma(yi = Z, 
                   sei = SE, 
                   mods = ~I(SE^2), 
                   data=datMA,
                   method = "FE")
  }
  if (error == "multiplicative") {
    peeseOut = lm(Z ~ I(SE^2), 
                  weights = 1/SE,
                  data=datMA)
  }
  return(peeseOut)
}

PEESE()
```
PEESE finds a significant effect



# funnel plot with PET line and conditional PEESE line ----
```{r}

funnelPETPEESE = function(dataset, 
                          error = "additive",
                          alwaysPEESE=T, plotName=NULL, printText = T,
                          ...) {
  naiveModel = naive(dataset)
  petModel = PET(dataset, error)
  peeseModel = PEESE(dataset, error)
  # make funnel plot
  funnel(naiveModel, ...)
  if (printText == T) title(plotName, line=3)
  if (printText == F) title(plotName)
  if (error == "additive") {
    if (printText == T) naiveModel$b[1] %>% 
      tanh %>% 
      round(3) %>%
      paste("Naive meta estimate, r =", .) %>%
      mtext(side=1)
    # add line and text from PET
    petModel %$% 
      abline(a = -b[1]/b[2], b = 1/b[2])
    r = petModel$b[1] %>% tanh %>% round(3)
    p.effect = petModel$pval[1] %>% round(3)
    p.bias = petModel$pval[2] %>% round(3)
    if (printText == T) mtext(paste("PET r = ", r
                                    , ", p-effect = ", p.effect
                                    , ", p-bias = ", p.bias
                                    , sep=""))
    points(x = petModel$b[1], y=0, cex=1.5)
    #abline(v = petModel$b[1], lty = 2)
    #lines(x = rep(petModel$b[1], 2), y = c(ymin, ymin - .1)
    # add line and text from PEESE
    if(petModel$pval[1] < .05 || alwaysPEESE == T) {
      grid = 
        naiveModel$vi %>%
        raise_to_power(.5) %>%
        max %>%
        seq(0, ., .001) %>%
        data.frame("Std.Err" = .)
      grid$Var = grid$Std.Err^2
      grid$Fisher.s.Z = 
        peeseModel$b[1] + peeseModel$b[2]*grid$Var
      grid %$% lines(x=Fisher.s.Z, y=Std.Err, typ='l')
      points(x = (peeseModel$b[1]), y=0, cex=1.5, pch=5)
      #abline(v = peeseModel$b[1], lty = 2)
      if (printText == T) { 
        peeseModel$b[1] %>%
          tanh %>%
          round(3) %>%
          paste("PEESE r =", .) %>%
          mtext(line = 1)
      }
    }
  }
  if(error == "multiplicative") {
    if (printText == T) naiveModel$b[1] %>% 
      tanh %>% 
      round(3) %>%
      paste("Naive meta estimate, r =", .) %>%
      mtext(side=1)
    # add line and text from PET
    b = summary(petModel)$coefficients[,1]
    petModel %$% 
      abline(a = -b[1]/b[2], b = 1/b[2])
    r = b[1] %>% tanh %>% round(3)
    p.effect = summary(petModel)$coefficients[1,4] %>% round(3)
    p.bias = summary(petModel)$coefficients[2,4] %>% round(3)
    if (printText == T) mtext(paste("PET r = ", r
                                    , ", p-effect = ", p.effect
                                    , ", p-bias = ", p.bias
                                    , sep=""))
    points(x = b[1], y=0, cex=1.5)
    #abline(v = b[1], lty = 2)
    # add line and text from PEESE
    if(summary(petModel)$coefficients[1,4] < .05 || alwaysPEESE == T) {
      grid = 
        naiveModel$vi %>%
        raise_to_power(.5) %>%
        max %>%
        seq(0, ., .001) %>%
        data.frame("Std.Err" = .)
      grid$Var = grid$Std.Err^2
      b = summary(peeseModel)$coefficients[,1]
      grid$Fisher.s.Z = 
        b[1] + b[2]*grid$Var
      grid %$% lines(x=Fisher.s.Z, y=Std.Err, typ='l')
      points(x = b[1], y=0, cex=1.5, pch=5)
      #abline(v = b[1], lty = 2)
      if (printText == T) {
        b[1] %>%
          tanh %>%
          round(3) %>%
          paste("PEESE r =", .) %>%
          mtext(line = 1)
      }
    }
  }
}
funnelPETPEESE()
```
```{r}
dpi = 150

png(filename = "FixedEffectsPETPEESE.png", width = (500 * (dpi/72)), height = (300 * (dpi/72)), res = dpi, bg = "transparent")

funnelPETPEESE()

dev.off()
```
# PET PEESE with Random Effects Model

# naive meta-analysis ----
```{r}


?rma

naive2 = function(dataset, ...) {
  rma(yi = Z,
      sei = SE,
      data = datMA,
      method = "REML",
      ...)
}
naive2()
```
Naive random effects model finds significant effect -1.25

# Basic PET
```{r}
PET2=function(dataset, error = "additive") {
  if (error == "additive") {
    petOut = rma(yi = Z, 
                 sei = SE, 
                 mods = ~SE, 
                 data=datMA,
                 method = "REML")
  }
  if (error == "multiplicative") {
    petOut = lm(Z ~ SE,
                weights = 1/SE,
                data=datMA)
  }
  return(petOut)
}
PET2()
```
Random effects generates a non-significant PET results (therefore should stop here) + CI that crosses zero & very large effect.


# Basic PEESE Function
```{r}

PEESE2=function(dataset, error = "additive") {
  if (error == "additive") {
    peeseOut = rma(yi = Z, 
                   sei = SE, 
                   mods = ~I(SE^2), 
                   data=datMA,
                   method = "REML")
  }
  if (error == "multiplicative") {
    peeseOut = lm(Z ~ I(SE^2), 
                  weights = 1/SE,
                  data=datMA)
  }
  return(peeseOut)
}

PEESE2()
```
Also NS, CI crossing zero.

# funnel plot with PET line and conditional PEESE line ----
```{r}

funnelPETPEESE2 = function(dataset, 
                          error = "additive",
                          alwaysPEESE=T, plotName=NULL, printText = T,
                          ...) {
  naiveModel = naive(dataset)
  petModel = PET2(dataset, error)
  peeseModel = PEESE2(dataset, error)
  # make funnel plot
  funnel(naiveModel, ...)
  if (printText == T) title(plotName, line=3)
  if (printText == F) title(plotName)
  if (error == "additive") {
    if (printText == T) naiveModel$b[1] %>% 
      tanh %>% 
      round(3) %>%
      paste("Naive meta estimate, r =", .) %>%
      mtext(side=1)
    # add line and text from PET
    petModel %$% 
      abline(a = -b[1]/b[2], b = 1/b[2])
    r = petModel$b[1] %>% tanh %>% round(3)
    p.effect = petModel$pval[1] %>% round(3)
    p.bias = petModel$pval[2] %>% round(3)
    if (printText == T) mtext(paste("PET r = ", r
                                    , ", p-effect = ", p.effect
                                    , ", p-bias = ", p.bias
                                    , sep=""))
    points(x = petModel$b[1], y=0, cex=1.5)
    #abline(v = petModel$b[1], lty = 2)
    #lines(x = rep(petModel$b[1], 2), y = c(ymin, ymin - .1)
    # add line and text from PEESE
    if(petModel$pval[1] < .05 || alwaysPEESE == T) {
      grid = 
        naiveModel$vi %>%
        raise_to_power(.5) %>%
        max %>%
        seq(0, ., .001) %>%
        data.frame("Std.Err" = .)
      grid$Var = grid$Std.Err^2
      grid$Fisher.s.Z = 
        peeseModel$b[1] + peeseModel$b[2]*grid$Var
      grid %$% lines(x=Fisher.s.Z, y=Std.Err, typ='l')
      points(x = (peeseModel$b[1]), y=0, cex=1.5, pch=5)
      #abline(v = peeseModel$b[1], lty = 2)
      if (printText == T) { 
        peeseModel$b[1] %>%
          tanh %>%
          round(3) %>%
          paste("PEESE r =", .) %>%
          mtext(line = 1)
      }
    }
  }
  if(error == "multiplicative") {
    if (printText == T) naiveModel$b[1] %>% 
      tanh %>% 
      round(3) %>%
      paste("Naive meta estimate, r =", .) %>%
      mtext(side=1)
    # add line and text from PET
    b = summary(petModel)$coefficients[,1]
    petModel %$% 
      abline(a = -b[1]/b[2], b = 1/b[2])
    r = b[1] %>% tanh %>% round(3)
    p.effect = summary(petModel)$coefficients[1,4] %>% round(3)
    p.bias = summary(petModel)$coefficients[2,4] %>% round(3)
    if (printText == T) mtext(paste("PET r = ", r
                                    , ", p-effect = ", p.effect
                                    , ", p-bias = ", p.bias
                                    , sep=""))
    points(x = b[1], y=0, cex=1.5)
    #abline(v = b[1], lty = 2)
    # add line and text from PEESE
    if(summary(petModel)$coefficients[1,4] < .05 || alwaysPEESE == T) {
      grid = 
        naiveModel$vi %>%
        raise_to_power(.5) %>%
        max %>%
        seq(0, ., .001) %>%
        data.frame("Std.Err" = .)
      grid$Var = grid$Std.Err^2
      b = summary(peeseModel)$coefficients[,1]
      grid$Fisher.s.Z = 
        b[1] + b[2]*grid$Var
      grid %$% lines(x=Fisher.s.Z, y=Std.Err, typ='l')
      points(x = b[1], y=0, cex=1.5, pch=5)
      #abline(v = b[1], lty = 2)
      if (printText == T) {
        b[1] %>%
          tanh %>%
          round(3) %>%
          paste("PEESE r =", .) %>%
          mtext(line = 1)
      }
    }
  }
}
funnelPETPEESE2()
```

```{r}

dpi = 150

png(filename = "RandomEffectsPETPEESE.png", width = (500 * (dpi/72)), height = (300 * (dpi/72)), res = dpi, bg = "transparent")

funnelPETPEESE2()

dev.off()
```
```{r}

```

