## Setup

Firstly, let's subset the dataset selecting our predictor (i.e., the **debulking**), the dataset name and the selected gene.

```{r}
datl <- readRDS(here("03-biostatistics", "data", "gene_data_clean.rds"))
dat <- dplyr::bind_rows(datl, .id = "dataset")

# creating the binary debulking variable
dat$debulking01 <- ifelse(dat$debulking == "optimal", 1, 0)

gene <- "ABCA1"
dat_meta <- dat[, c("dataset", "debulking", gene)]
head(dat_meta)
```

## Prevalence

Now let's see which is the prevalence of *optimal* debulking i.e. $S_1$

::: {.panel-tabset}

### Code

```{r}
prev_and_n <- function(data){
  data.frame(
    p = mean(data$debulking01, na.rm = TRUE),
    n = nrow(data)
  )
}
prev <- lapply(split(dat_meta, dat_meta$dataset), prev_and_n)
prev <- dplyr::bind_rows(prev, .id = "dataset")
head(prev)
```

### Plot

```{r}
#| code-fold: true
prev |> 
  ggplot(aes(x = p*100, y = dataset)) +
  geom_col(width = 0.2) +
  xlab("Prevalence (%)") +
  xlim(c(0, 100))
```

:::

## debulking ~ gene

Let's see the density plots for dataset with the gene `r gene`:

```{r}
#| code-fold: true
dat_meta |> 
  drop_na(debulking) |> 
  ggplot(aes(x = ABCA1, fill = debulking)) +
  geom_density() +
  facet_wrap(~dataset) +
  theme(legend.position = c(0.9,0.1),
        legend.title = element_blank(),
        axis.title.y = element_blank())
```

In some datasets, the gene `r gene` discriminate more between the the debulking outcomes.

## AUC and standard error

::: {.panel-tabset}

### ROC

```{r}
roc_list <- lapply(split(dat_meta, dat_meta$dataset), 
                   function(x) pROC::roc(debulking ~ ABCA1, data = x))
```

### Analytical method

```{r}
aucs <- sapply(roc_list,  auc)
# the var function is used to compute the sampling variability
aucs_se <- sapply(roc_list, function(x) sqrt(var(x)))
auc_d <- data.frame(dataset = names(aucs), auc = aucs, se = aucs_se)
ci <- sapply(roc_list, function(x) pROC::ci.auc(x))
auc_d$lower <- ci[1, ]
auc_d$upper <- ci[3, ]

head(auc_d)
```

### Plot

```{r}
#| code-fold: true
auc_d |> 
  ggplot(aes(x = auc, y = dataset)) +
  geom_point() +
  geom_segment(aes(x = lower, xend = upper, y = dataset, yend = dataset)) +
  xlim(c(0, 1))
```

### Bootstrapping

```{r}
boot_auc <- function(data, direction = "auto", nboot = 10){
  aucs <- rep(0, nboot)
  for(i in 1:nboot){
    idx <- sample(1:nrow(data), nrow(data), replace = TRUE)
    datb <- data[idx, ]
    aucs[[i]] <- pROC::auc(datb[, 2] ~ datb[, 3], direction = direction, quiet = TRUE)[[1]]
  }
  return(aucs)
}

# get the direction, auto detecting direction could lead to biased AUC see https://cran.r-project.org/web/packages/pROC/pROC.pdf
direction <- sapply(roc_list, function(x) x$direction)

auc_list <- mapply(function(d, dir) boot_auc(d, dir), split(dat_meta, dat_meta$dataset), direction)
aucs_m <-  apply(auc_list, 2, median)
aucs_se <- apply(auc_list, 2, sd)
aucs_ci <- apply(auc_list, 2, function(x) quantile(x, c(0.025, 0.975)))
aucs_d <- data.frame(auc = aucs_m, se = aucs_se, lower = aucs_ci[1, ], upper = aucs_ci[2, ], dataset = names(roc_list))

head(aucs_d)
```

### Plot

```{r}
#| code-fold: true
aucs_d |> 
  ggplot(aes(x = auc, y = dataset)) +
  geom_point() +
  geom_segment(aes(x = lower, xend = upper, y = dataset, yend = dataset)) +
  xlim(c(0, 1))
```

:::

## Meta-analysis

```{r}
library(metafor)
auc_d$var <- auc_d$se^2

fit_rma <- rma(auc, var, data = auc_d)
summary(fit_rma)
```

## Meta-analysis

We need to adjust the test statistics because the null value with the AUC is not 0 but 0.5.

$$
z = \frac{\hat{AUC} - AUC_0}{\hat{SE_{AUC}}}
$$

$$
z = \frac{\hat{AUC} - 0.5}{SE_{\hat{AUC}}}
$$

```{r}
# pvalue testing the null of 0.5
pnorm((fit_rma$b[[1]] - 0.5)/fit_rma$se, lower.tail = FALSE)*2
```

## Forest Plot

```{r}
forest(fit_rma, refline = 0.5)
```

# Comments? {.question}